---
title: "references"
date: "`r doc_date()`"
vignette: >
  %\VignetteIndexEntry{References for metilclock using Bioconductor's ExperimentHub}
  %\VignetteEngine{knitr::rmarkdown}
output: 
  BiocStyle::html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

The methylclockData package is a repository of a few public datasets that
needs the *methylclock* package to estimate chronological and gestational DNA 
methylation (DNAm) age as well as biological age using different methylation clocks.

We use the data from package `meffil` (@min2018meffil) that provides some
functions to estimate cell counts using predefined datasets. 

# Cell counts
The EEAA method implemented in [methylclock](https://github.com/isglobal-brge/methylclock) requires to estimate cell counts. We use the package `meffil` (@min2018meffil) that provides some functions to estimate cell counts using predefined datasets. So far, the following data is available:

- **"blood gse35069 complete"**: methylation profiles from @reinius2012differential for purified blood cell types. It includes CD4T, CD8T, Mono, Bcell, NK, Neu and Eos.
- **"blood gse35069"**: methylation profiles from @reinius2012differential for purified blood cell types. It includes CD4T, CD8T, Mono, Bcell, NK and Gran.
- **"blood gse35069 chen"**: methylation profiles from @chen2017epigenome blood cell types. It includes CD4T, CD8T, Mono, Bcell, NK, Neu and Eos.
- **"andrews and bakulski cord blood"**. Cord blood reference from @bakulski2016dna. It includes Bcell, CD4T, CD8T, Gran, Mono, NK and nRBC.
- **"cord blood gse68456"** Cord blood methylation profiles from @de2015nucleated. It includes CD4T, CD8T, Mono, Bcell, NK, Neu, Eos and RBC.
- **"gervin and lyle cord blood"** Cord blood reference generated by Kristina Gervin and Robert Lyle, available at `miffil` package. It includes CD14, Bcell, CD4T, CD8T, NK, Gran.
- **"saliva gse48472"**: Reference generated from the multi-tissue pannel from @slieker2013identification. It includes Buccal, CD4T, CD8T, Mono, Bcell, NK, Gran.

# How to load data

In the below example, we show how one can download this dataset from 
ExperimentHub.

```{r get-tcga}
library(ExperimentHub)
eh <-  ExperimentHub()
query(eh , "methylclockData")
```
For more information in how loading and use of the data, please, refer to [`MethylClock` vignette](https://github.com/isglobal-brge/methylclock) 


# sessionInfo()

```{r sessionInfo}
sessionInfo()
```
